### 2015-01-21 doc/thesis/typedlua.tex
Improving the text about table types, aliases, and modules

\begin{figure}[!ht]
\textbf{Types}\\
\dstart
\begin{align*}
\textit{type} ::= & \;\; \textit{primarytype} \; [\texttt{`?'}]\\
\textit{primarytype} ::= & \;\; \textit{literaltype} \; | \;
  \textit{basetype} \; | \;
  \textbf{nil} \; | \;
  \textbf{value} \; | \;
  \textbf{any} \; | \;
  \textbf{self} \; | \;
  \textit{Name}\\
& | \; \textit{functiontype} \; | \;
  \textit{tabletype} \; | \;
  \textit{primarytype} \; \texttt{`|'} \; \textit{primarytype}\\
\textit{literaltype} ::= & \;\; \textbf{false} \; | \;
  \textbf{true} \; | \;
  \textit{Int} \; | \;
  \textit{Float} \; | \;
  \textit{String}\\
\textit{basetype} ::= & \;\; \textbf{boolean} \; | \;
  \textbf{integer} \; | \;
  \textbf{number} \; | \;
  \textbf{string}\\
\textit{functiontype} ::= & \;\; \textit{tupletype} \; \texttt{`->'} \; \textit{rettype}\\
\textit{tupletype} ::= & \;\; \texttt{`('} \; [typelist] \; \texttt{`)'}\\
\textit{typelist} ::= & \;\; \textit{type} \; \{\texttt{`,'} \; \textit{type}\} \; [\texttt{`*'}]\\
\textit{rettype} ::= & \;\; \textit{type} \; | \;
  \textit{uniontuple} \; [\texttt{`?'}]\\
\textit{uniontuple} ::= & \;\; \textit{tupletype} \; | \;
  \textit{uniontuple} \; \texttt{`|'} \; \textit{uniontuple}\\
\textit{tabletype} ::= & \;\; \texttt{`\{'} \; [\textit{tabletypebody}] \; \texttt{`\}'}\\
\textit{tabletypebody} ::= & \;\; \textit{hashtype} \; | \;
  \textit{recordtype}\\
\textit{hashtype} ::= & \;\; [\textit{keytype} \; \texttt{`:'}] \; \textit{type}\\
\textit{keytype} ::= & \;\; \textit{basetype} \; | \;
  \textbf{value}\\
\textit{recordtype} ::= & \;\; \textit{recordfield} \; \{\texttt{`,'} \; \textit{recordfield}\} \; [\texttt{`,'} \; \textit{type}]\\
\textit{recordfield} ::= & \;\; [\textbf{const}] \; \textit{literaltype} \; \texttt{`:'} \; \textit{type}
\end{align*}
\dend
\caption{The concrete syntax of Typed Lua types}
\label{fig:types}
\end{figure}

@@

Typed Lua allows the assignment \texttt{x = x or e} to refine
the type of \texttt{x} from \texttt{t|nil} to \texttt{t}
if it matches the following rules: 
the type of \texttt{e} is a subtype of \texttt{t},
and the variable \texttt{x} is local to the current function
and it is not being assigned in another function.
The refinement only affects the type of \texttt{x} until the
next assignment to \texttt{x}.
In the case of \texttt{greeting}, the assignment refines its type
from \texttt{string|nil} to \texttt{string}.

@@

Typed Lua has a small set of \texttt{type} predicates that allows
programmers to constrain the type of a local variable inside a condition.
This example uses the \texttt{type(x) == "string"} predicate that
constrains the type of \texttt{x} from \texttt{string|t} to
\texttt{string} inside the \texttt{if} body and to \texttt{t}
inside the \texttt{else} body. 
This is a simplified form of \emph{flow typing} \citep{guha2011tlc,tobin-hochstadt2010ltu}.
As with \texttt{or}, the variable must be local to the function
and it is not being assigned in another function.

@@

Tables are the only mechanism that Lua has to build data structures;
they are associative arrays where any value (except \texttt{nil})
can be used as a key.
Programmers can use tables to represent tuples, arrays (dense or sparse),
records, graphs, modules, objects, etc.
Lua has syntactic sugar for indexing tables as records:
\texttt{t.k} is syntactic sugar for \texttt{t["k"]}, and means that
we are accessing the field \texttt{k} from table \texttt{t}. 
In this section, we show how Typed Lua encodes tables as maps,
arrays, and records.

Table types are the only framework that Typed Lua has to represent
the different uses that a Lua table has.
A table type $\texttt{\{k}{_\texttt{1}}\texttt{:t}_{\texttt{1}}
\texttt{, ..., k}{_\texttt{n}}\texttt{:t}_{\texttt{n}}\texttt{\}}$ represents a map
from values of type $\texttt{k}_{\texttt{i}}$ to values of type
$\texttt{t}_{\texttt{i}}$.

\begin{figure}[!ht]
\textbf{Table types}\\
\dstart
\begin{align*}
\textit{tabletype} ::= & \;\; \texttt{`\{'} \; [\textit{tabletypebody}] \; \texttt{`\}'}\\
\textit{tabletypebody} ::= & \;\; [\textit{keytype} \; \texttt{`:'}] \; \textit{type} \; | \;
  \textit{recordtype}\\
\textit{keytype} ::= & \;\; \textit{basetype} \; | \;
  \textbf{value}\\
\textit{recordtype} ::= & \;\; \textit{recordfield} \; \{\texttt{`,'} \; \textit{recordfield}\} \; [\texttt{`,'} \; \textit{type}]\\
\textit{recordfield} ::= & \;\; [\textbf{const}] \; \textit{literaltype} \; \texttt{`:'} \; \textit{type}
\end{align*}
\dend
\caption{The concrete syntax of Typed Lua table types}
\label{fig:tables}
\end{figure}

Although Typed Lua has just one framework to represent tables,
it has three different syntaxes for defining the type of a table:
\begin{align}
\label{syn:table1}
\texttt{`\{'} \; \textit{keytype} \; \texttt{`:'} \; \textit{type} \; \texttt{`\}'}\\
\label{syn:table2}
\texttt{`\{'} \; \textit{type} \; \texttt{`\}'}\\
\label{syn:table3}
\texttt{`\{'} \; \textit{literaltype} \; \texttt{`:'} \; \textit{type} \;
\{\texttt{`,'} \; \textit{literaltype} \; \texttt{`:'} \; \textit{type}\} \;
[\texttt{`,'} \; \textit{type}] \; \texttt{`\}'}
\end{align}
The syntax \ref{syn:table1} defines table types for maps:
it is written \texttt{\{k:t\}},
and maps to the table type \texttt{\{k:t|nil\}}.
This table type represents a map that maps values of type
\texttt{k} to values of type \texttt{t|nil},
as Lua returns \texttt{nil} when we use a non-existing key to index a table.
The syntax \ref{syn:table2} defines table types for arrays:
it is written \texttt{\{t\}},
and maps to the table type \texttt{\{integer:t|nil\}}.
This table type is just a syntactic sugar to the definition of a
map that maps values of type \texttt{integer} to values of type \texttt{t|nil}.
The syntax \ref{syn:table3} defines table types for records with an
optional array part: it is written
$\texttt{\{l}{_\texttt{1}}\texttt{:t}_{\texttt{1}}
\texttt{, ..., l}{_\texttt{n}}\texttt{:t}_{\texttt{n}}\texttt{\}}$,
and maps to the same table type.
This table type represents a record that maps the literal types
$\texttt{l}_{\texttt{i}}$\texttt{, ..., }$\texttt{l}_{\texttt{n}}$
to values of types $\texttt{t}_{\texttt{i}}$\texttt{, ..., }$\texttt{t}_{\texttt{n}}$.

Next, we show one example of table type to type a map from strings to integers:
\begin{verbatim}
    local t:{string:integer} = { foo = 1 } 
    local x:integer = t.foo               -- compile-time error
    local y:integer? = t.bar              -- y gets nil
    local z:integer = t["bar"] or 0       -- z gets 0 
\end{verbatim}

The second line of this example gives a type error because we are
attempting to assign a value of type \texttt{integer|nil} to a
variable that accepts only values of type \texttt{integer}.
Although the field \texttt{bar} does not exist in \texttt{t}, the third
line of this example does not give a type error because the
annotated type matches the type of the values that can be stored in
\texttt{t}.
The last line shows that the \texttt{or} idiom is also useful to
give a default value to a missing table field.

Next, we show one example of table type to type an array:
\begin{verbatim}
    local days:{string} = { "Sunday", "Monday", "Tuesday",
                            "Wednesday", "Thursday", "Friday",
                            "Saturday" }
    local x = days[1]                    -- x gets "Sunday"
    local y = days[8]                    -- y gets nil
\end{verbatim}

In this example, the type of both variables \texttt{x} and \texttt{y}
is \texttt{string|nil}.
An inconvenient aspect of making the types of maps and arrays always
include the type \texttt{nil} is to overload the programmers,
as they need to use \texttt{or} or \texttt{if} to narrow the type of
the elements they are accessing.

When we know that a list has fixed elements, we can leave the
variable declaration unannotated and let local type inference assign
a more specific table type to the variable.
If we remove the annotation in this example, the compiler uses the
syntax of records to infer the following table type to \texttt{days}:
\texttt{\{1:string, 2:string, 3:string, 4:string, 5:string, 6:string, 7:string\}}.

When we use the inferred table type while initializing a table, the
compiler gives an error message when we try to access an index that
is out of bounds.

We can also use the style of the inferred type to define heterogeneous
tuples:
\begin{verbatim}
    local album:{1:string, 2:integer, 3:string} =
      { "Transformer", 1972, "Lou Reed" }
\end{verbatim}

The next example shows how we can annotate a \texttt{map} function
that receives another function as its parameter and uses this
function to alter the elements of a list of numbers, which is
also a parameter of \texttt{map}:
\begin{verbatim}
    local function map (f:(number)->(number), l:{number})
      for i = 1, #l do
        l[i] = f(l[i] or 0)
      end
    end
\end{verbatim}

Next, we show one example of table type to type a record:
\begin{verbatim}
    local person:{"firstname":string, "lastname":string} =
      { firstname = "Lou", lastname = "Reed" } 
\end{verbatim}

Type inference would infer the very same table type that we used to
annotate the variable \texttt{person}.

Lua programmers often build records incrementally, that is,
they usually declares a local variable with an empty table,
and then use assignment to add fields to this table:
\begin{verbatim}
    local person = {}
    person.firstname = "Lou"
    person.lastname = "Reed"
    print("bye bye " .. person.firstname)  -- bye bye Lou
\end{verbatim}

In this example, we want to refine the type of \texttt{person} as
we build the table:
starting with \texttt{\{\}}, and then refining to \texttt{\{"firstname":string\}},
and finally reaching \texttt{\{"firstname":string, "lastname":string\}}.
This type change is trickier than the one that we introduced for narrowing
union types, as we are not just allowing the programmer to change
the type of the variable \texttt{person}, but we are actually allowing
the programmer to change the type of the value that \texttt{person} points to.
This type change is safe in this example, but it is not in the following one:
\begin{verbatim}
    local person = {}
    local bogus = person
    person.firstname = "Lou"
    person.lastname = "Reed"
    bogus.firstname = true               -- compile-time error
    print("bye bye " .. person.firstname)
\end{verbatim}

In this example, the fifth line is unsound because it allows the
programmer to change the type of the value that is stored in
\texttt{person.firstname}, and makes the last line crash during
run-time.
On the first line, we assign an empty table to \texttt{person}.
On the second line, we assign the type of \texttt{person} to
\texttt{bogus}.
On the third and fourth lines we change the type of \texttt{person}
as we did in the previous example.
On the fifth line we try to change the type of \texttt{bogus}
from \texttt{\{\}} to \texttt{\{"firstname":boolean\}},
but if we allow this type change we also allow changing the value
that is stored in \texttt{person.firstname}, regardless of its type.
Taking the changes individually looks fine, but the truth is that
aliasing makes one of them unsound.

The location of the change also matters, as the next example shows:
\begin{verbatim}
    local person = { lastname = "Reed" }
    local function spoil ()
      person.firstname = nil             -- compile-time error
    end
    person.firstname = "Lou"
    spoil()
    print("bye bye " .. person.firstname)
\end{verbatim}

In this example, the third line is unsound because it allows the
programmer to change the type of \texttt{person} outside of the scope
that it was declared.
The call to \texttt{spoil} erases the field \texttt{firstname} from
\texttt{person}, and makes the last line crash during run-time.

It is also not sound to allow the type of the fields to change,
as the next example shows:
\begin{verbatim}
    local person = { firstname = "Lewis", lastname = "Reed" }
    person.middlename = "Allan"
    person.firstname = 1942              -- compile-time error
    person.lastname = 2013               -- compile-time error
    print("bye bye " .. person.firstname)
\end{verbatim}

In this example, both third and fourth line are unsound because
they are trying to change the type of fields that already exist.

Typed Lua tags a variable that holds a table type as either
\emph{open} or \emph{closed}.
If a variable gets its type from a table constructor then it
is open, otherwise it is always closed.

Typed Lua uses three restrictions to decide whether it is
sound to allow field assignment to change the type of an
open variable or not:
the variable must be local to the current block,
the new type must be a subtype of the old one, and
the variable cannot have been assigned in another function.

We can create aliases to open variables, but they are always closed.
Any mutation on these aliases is not a problem, as the type of
the original reference can only change towards a subtype.
For mutable fields this means that the type of the field cannot
change after it is added to the type of the table.

Table types along with their refinement through field assignment
are the foundation for defining modules and objects in Typed Lua.
In Section \ref{sec:modules} we show that the refinement of table
types through field assignment is crucial for making Typed Lua
support idiomatic definition of Lua modules.
In Section \ref{sec:oop} we show that the refinement of table
types is also helpful for defining classes and objects.

@@

Typed Lua includes \emph{interfaces} as syntactic sugar to table types
that specify records, as writing table types can be unwieldy when
records get bigger and the types of record fields get more complicated.
For instance, we can use the following interface as an alias to the
type of the record we showed in the previous section:
\begin{verbatim}
    local interface Person
      firstname:string
      lastname:string
    end
\end{verbatim}

\begin{figure}[!ht]
\textbf{Type alias}\\
\dstart
\begin{align*}
\textit{typealias} ::= & \; \textbf{typealias} \; \textit{Name} \; \texttt{`='} \; \textit{type}
\end{align*}
\dend
\caption{The concrete syntax of Typed Lua type aliases}
\label{fig:aliases}
\end{figure}

\begin{figure}[!ht]
\textbf{Interfaces}\\
\dstart
\begin{align*}
\textit{interface} ::= & \; \textbf{interface} \; \textit{typedec}\\
\textit{typedec} ::= & \; \textit{Name} \; { \; \textit{decitem} \; } \; \textbf{end}\\
\textit{decitem} ::= & \; \textit{idlist} \; \texttt{`:'} \; \textit{idtype}\\
\textit{idtype} ::= & \; \textit{type} \; | \; \textit{methodtype}\\
\textit{idlist} ::= & \; \textit{id} \; \{ \; \texttt{`,'} \; \textit{id} \; \}\\
\textit{id} ::= & \; [\textbf{const}] \; \textit{Name}\\
\textit{typealias} ::= & \; \textbf{typealias} \; \textit{Name} \; \texttt{`='} \; \textit{type}
\end{align*}
\dend
\caption{The concrete syntax of Typed Lua interfaces}
\label{fig:interfaces}
\end{figure}

The declaration of the interface \texttt{Person} is an alias to the
table type \texttt{\{"firstname":string, "lastname":string\}} in the
remainder of the current scope.
Now we can use \texttt{Person} in type declarations:
\begin{verbatim}
    local function byebye (person:Person):string
      return "Goodbye " .. person.firstname ..
             " " .. person.lastname
    end

@@

    print(greeter(user1)) -- Goodbye Lewis Reed 
    print(greeter(user2)) -- compile-time error
    print(greeter(user3)) -- Goodbye Lou Reed
    print(greeter(user4)) -- compile-time error
\end{verbatim}

@@

If our record type has optional fields, that is, fields that can be \texttt{nil},
we need to use an explicit type declaration when declaring a
variable of this record type, as we show in the following example:
\begin{verbatim}
    local interface Person
      firstname:string
      middlename:string?
      lastname:string
    end
    local user1:Person = { firstname = "Lewis",
                           middlename = "Allan",
                           lastname = "Reed" }
    local user2:Person = { lastname = "Reed",
                           firstname = "Lou" }
\end{verbatim}

We need an explicit type declaration because neither of the types
that the Typed Lua compiler assigns to the above table constructors
is a subtype of
\texttt{\{"firstname":string, "middlename":string|nil, "lastname":string\}},
the type that \texttt{Person} describes.

We can also use interfaces to define recursive types such as liked lists:
\begin{verbatim}
    local interface Element
      info:integer
      next:Element?
    end
\end{verbatim}

We need to explicitly annotate \texttt{next} being of type
\texttt{Element|nil} because in our type system \texttt{nil}
is not the bottom type, that is, \texttt{nil} is subtype of itself only.
Typed Lua has this behavior because in Lua \texttt{nil} represents
the absence of an useful value.
Thus, making \texttt{nil} the bottom type would severally reduce the amount
of type safety in the system.

@@

    while l do
      if l then print(l.info) end
      l = l.next                  -- compile-time error
    end
\end{verbatim}

However, this example has one compile-time error because we
are attempting to index the variable \texttt{l},
which can be \texttt{nil}.
Notice that we would have the same compile-time error while
printing the list element if we had not narrowed the type of
\texttt{l}.
To eliminate this compile-time error we have to write a
function that narrows the type of \texttt{l} before using it:
\begin{verbatim}
    local function get_next (l:Element?)
      if l then return l.next else return nil end
    end
\end{verbatim}

Now we can use \texttt{get\string_next} to rewrite the \texttt{while}
from the previous example, thus eliminating the compile-time error:
\begin{verbatim}
    while l do
      if l then print(l.info) end
      l = get_next(l)
    end
\end{verbatim}

Typed Lua also includes a way to create aliases for types that does not
follow the records syntax.
However, we limit aliases to union types, function types, table types,
and recursive types.
The reason for this restriction is to avoid the creation of aliases
to predefined types.

For instance, we can use the \texttt{typealias} construct to define an
alias to the type of the abstract syntax tree of a language of simple
arithmetic expressions:
\begin{verbatim}
    local typealias Exp = {"tag":"Number", 1:number}
                        | {"tag":"Add", 1:Exp, 2:Exp}
                        | {"tag":"Sub", 1:Exp, 2:Exp}
                        | {"tag":"Mul", 1:Exp, 2:Exp}
                        | {"tag":"Div", 1:Exp, 2:Exp}
\end{verbatim}

>>>>

\begin{figure}[!ht]
\textbf{Types}\\
\dstart
\begin{align*}
\textit{type} ::= & \;\; \textit{primarytype} \; [\texttt{`?'}]\\
\textit{primarytype} ::= & \;\; \textit{literaltype} \; | \;
  \textit{basetype} \; | \;
  \textbf{nil} \; | \;
  \textbf{value} \; | \;
  \textbf{any} \; | \;
  \textbf{self} \; | \;
  \textit{Name}\\
& | \; \textit{functiontype} \; | \;
  \textit{tabletype} \; | \;
  \textit{primarytype} \; \texttt{`|'} \; \textit{primarytype}\\
\textit{literaltype} ::= & \;\; \textbf{false} \; | \;
  \textbf{true} \; | \;
  \textit{Int} \; | \;
  \textit{Float} \; | \;
  \textit{String}\\
\textit{basetype} ::= & \;\; \textbf{boolean} \; | \;
  \textbf{integer} \; | \;
  \textbf{number} \; | \;
  \textbf{string}\\
\textit{functiontype} ::= & \;\; \textit{tupletype} \; \texttt{`->'} \; \textit{rettype}\\
\textit{tupletype} ::= & \;\; \texttt{`('} \; [typelist] \; \texttt{`)'}\\
\textit{typelist} ::= & \;\; \textit{type} \; \{\texttt{`,'} \; \textit{type}\} \; [\texttt{`*'}]\\
\textit{rettype} ::= & \;\; \textit{type} \; | \;
  \textit{uniontuple} \; [\texttt{`?'}]\\
\textit{uniontuple} ::= & \;\; \textit{tupletype} \; | \;
  \textit{uniontuple} \; \texttt{`|'} \; \textit{uniontuple}\\
\textit{tabletype} ::= & \;\; \texttt{`\{'} \; [\textit{tabletypebody}] \; \texttt{`\}'}\\
\textit{tabletypebody} ::= & \;\; \textit{maptype} \; | \;
  \textit{recordtype}\\
\textit{maptype} ::= & \;\; [\textit{keytype} \; \texttt{`:'}] \; \textit{type}\\
\textit{keytype} ::= & \;\; \textit{basetype} \; | \;
  \textbf{value}\\
\textit{recordtype} ::= & \;\; \textit{recordfield} \; \{\texttt{`,'} \; \textit{recordfield}\} \; [\texttt{`,'} \; \textit{type}]\\
\textit{recordfield} ::= & \;\; [\textbf{const}] \; \textit{literaltype} \; \texttt{`:'} \; \textit{type}
\end{align*}
\dend
\caption{The concrete syntax of Typed Lua types}
\label{fig:types}
\end{figure}

@@

Typed Lua allows the assignment \texttt{v = v or e} to refine
the type of \texttt{v} from \texttt{t|nil} to \texttt{t}
if it matches the following rules: 
the type of \texttt{e} is a subtype of \texttt{t},
and the variable \texttt{v} is local to the current function
and it is not being assigned in another function.
The refinement only affects the type of \texttt{v} until the
next assignment to \texttt{v}.
In the case of \texttt{greeting}, the assignment refines its type
from \texttt{string|nil} to \texttt{string}.

@@

Typed Lua has a small set of \texttt{type} predicates that allows
programmers to constrain the type of a local variable inside a condition.
This example uses the \texttt{type(v) == "string"} predicate that
constrains the type of \texttt{v} from \texttt{string|t} to
\texttt{string} inside the \texttt{if} body and to \texttt{t}
inside the \texttt{else} body. 
This is a simplified form of \emph{flow typing} \citep{guha2011tlc,tobin-hochstadt2010ltu}.
As with \texttt{or}, the variable must be local to the function
and it is not being assigned in another function.

@@

Tables are the only mechanism that Lua has to build data structures;
they are associative arrays where any value (except \texttt{nil})
can be used as a key.
Programmers can use tables to represent tuples, arrays (dense or sparse),
records, graphs, modules, objects, etc.
Lua has syntactic sugar for indexing tables as records:
\texttt{t.k} is syntactic sugar for \texttt{t["k"]}.
In this section, we show how Typed Lua types tables that encode maps,
arrays, and records.

\begin{figure}[!ht]
\textbf{Table types}\\
\dstart
\begin{align*}
\textit{tabletype} ::= & \;\; \texttt{`\{'} \; [\textit{tabletypebody}] \; \texttt{`\}'}\\
\textit{tabletypebody} ::= & \;\; \textit{maptype} \; | \;
  \textit{recordtype}\\
\textit{maptype} ::= & \;\; [\textit{keytype} \; \texttt{`:'}] \; \textit{type}\\
\textit{keytype} ::= & \;\; \textit{basetype} \; | \;
  \textbf{value}\\
\textit{recordtype} ::= & \;\; \textit{recordfield} \; \{\texttt{`,'} \; \textit{recordfield}\} \; [\texttt{`,'} \; \textit{type}]\\
\textit{recordfield} ::= & \;\; [\textbf{const}] \; \textit{literaltype} \; \texttt{`:'} \; \textit{type}
\end{align*}
\dend
\caption{The concrete syntax of Typed Lua table types}
\label{fig:tables}
\end{figure}

Figure \ref{fig:tables} shows that the concrete syntax of Typed Lua
table types is restricted to either the type of the empty table,
maps, arrays, or records with an optional array part.
We made this design choice due to the results that we obtained
about the usage of the table constructor, which we discussed
in Section \ref{sec:statistics}.
Later in this chapter we will show that we use the syntax of records
to type modules and objects.
We will also show that we need the \texttt{const} modifier while
typing object-oriented code.

The table type \texttt{\{k:t\}} represents the type of a map
from values of type \texttt{k} to values of type \texttt{t|nil}.
Table types that represent maps always include the type \texttt{nil},
because Lua returns \texttt{nil} when we use a non-existing key to
index a table. 
The types of the keys are restricted to either base types or the
type \texttt{value} due to the statistics of the table constructor,
which we discussed in Section \ref{sec:statistics}.
Next, we show one example of table type to type a map from strings to integers:
\begin{verbatim}
    local t:{string:integer} = { foo = 1 } 
    local x:integer = t.foo           --> compile-time warning
    local y:integer? = t.bar          --> y gets nil
    local z:integer = t["bar"] or 0   --> z gets 0 
\end{verbatim}

The second line of this example raises a warning, because we are
attempting to assign a value of type \texttt{integer|nil} to a
variable that accepts only values of type \texttt{integer}.
Although the field \texttt{bar} does not exist in \texttt{t}, the third
line of this example does not raise a warning, because the
annotated type matches the type of the values that can be stored in
\texttt{t}.
The last line shows that the \texttt{or} idiom is also useful to
give a default value to a missing table field.

The table type \texttt{\{t\}} represents the type of an array that 
maps values of type \texttt{integer} to values of type \texttt{t|nil}.
In other words, Typed Lua handles arrays as syntactic sugar to the
table type \texttt{\{integer:t\}}.
Next, we show one example of table type to type an array of strings:
\begin{verbatim}
    local days:{string} = { "Sunday", "Monday", "Tuesday",
                            "Wednesday", "Thursday", "Friday",
                            "Saturday" }
    local i = 5
    local x = days[1]                 --> x gets "Sunday"
    local y = days[8]                 --> y gets nil
    local z = days[i]                 --> z gets "Thursday"
\end{verbatim}

In this example, the type of \texttt{i} is \texttt{integer},
while the type of \texttt{x}, \texttt{y}, and \texttt{z} is
\texttt{string|nil}.
An inconvenient aspect of making the types of maps and arrays always
include the type \texttt{nil} is to overload the programmers,
as they need to use \texttt{or} or \texttt{if} to narrow the type of
the elements they are accessing.

The table type
$\texttt{\{l}{_\texttt{1}}\texttt{:t}_{\texttt{1}}
\texttt{, ..., l}{_\texttt{n}}\texttt{:t}_{\texttt{n}}\texttt{\}}$
represents the type of a record that maps literal types
$\texttt{l}_{\texttt{i}}$\texttt{, ..., }$\texttt{l}_{\texttt{n}}$
to values of types $\texttt{t}_{\texttt{i}}$\texttt{, ..., }$\texttt{t}_{\texttt{n}}$.
Most programming languages treat records as maps from names
to types, but we chose to use literal types instead of
names due to the statistics results that we obtained about
the usage of the table constructor, which we discussed in
Section \ref{sec:statistics}.
This means that we can use the syntax of records to type a
list that has fixed elements, like the variable \texttt{days} from
the previous example.

When we know that a list has fixed elements, we can leave the
variable declaration unannotated and let local type inference assign
a more specific table type to the variable.
If we remove the annotation in the previous example, the compiler uses the
syntax of records to infer the following table type to \texttt{days}:
\texttt{\{1:string, 2:string, 3:string, 4:string, 5:string, 6:string, 7:string\}}.

When we use the syntax of records to type an array, the compiler
raises a warning when we try to access an index that is out of bounds.
In the previous example, both \texttt{days[8]} and \texttt{days[i]}
raise warnings, as both the literal type \texttt{8} and the base type
\texttt{integer} do not map to any value.

We can also use the syntax of records to type heterogeneous tuples:
\begin{verbatim}
    local album:{1:string, 2:integer, 3:string} =
      { "Transformer", 1972, "Lou Reed" }
\end{verbatim}

Next, we show one example of table type to type a record that maps
names to strings:
\begin{verbatim}
    local person:{"firstname":string, "lastname":string} =
      { firstname = "Lou", lastname = "Reed" } 
\end{verbatim}

In these two previous examples, local type inference would infer the
very same table types that we used to annotate the variables
\texttt{album} and \texttt{person}.

Lua programmers often build records incrementally, that is,
they usually declare a local variable with an empty table,
and then use assignment to add fields to this table:
\begin{verbatim}
    local person = {}
    person.firstname = "Lou"
    person.lastname = "Reed"
    print("bye bye " .. person.firstname)  --> bye bye Lou
\end{verbatim}

In this example, we want to refine the type of \texttt{person} as
we build the table:
starting with \texttt{\{\}}, and then refining to \texttt{\{"firstname":string\}},
and finally reaching \texttt{\{"firstname":string, "lastname":string\}}.
This type change is trickier than the one that we introduced for narrowing
union types, as we are not just allowing the programmer to change
the type of the variable \texttt{person}, but we are actually allowing
the programmer to change the type of the value that \texttt{person} points to.

Typed Lua tags a variable that holds a table type as either
\emph{unique}, \emph{open}, or \emph{closed}.
If a variable gets is type from a table constructor then it is
unique, otherwise it is closed.
Any alias to unique variables makes them open, as we use the
tag open to keep track of unique variables that are aliased.
In the previous example, \texttt{person} has an unique type
because it has no alias.
Typed Lua also has different subtyping rules that handle these
different tags.
We explain these tags and their subtyping rules in more detail
in Chapter \ref{chap:system}.

Typed Lua uses three rules to decide whether it is sound to
allow field assignment to change the type of unique and open
variables:
the variable must be local to the current block,
the new type must only add new fields, and
the variable cannot have been assigned in another function.

The following example shows that it is not sound to allow the
type of the fields to change:
\begin{verbatim}
    local person = { firstname = "Lewis", lastname = "Reed" }
    person.middlename = "Allan"
    person.firstname = 1942           --> compile-time warning
    person.lastname = 2013            --> compile-time warning
    print("bye bye " .. person.firstname)
\end{verbatim}

In this example, both third and fourth lines are unsound because
they are trying to change the type of fields that already exist.

It is also not sound to allow changing the type of an alias:
\begin{verbatim}
    local person = {}
    local bogus = person
    person.firstname = "Lou"
    person.lastname = "Reed"
    bogus.firstname = true            --> compile-time warning
    print("bye bye " .. person.firstname)
\end{verbatim}

In this example, the fifth line is unsound because it allows the
programmer to change the type of the value that is stored in
\texttt{person.firstname}, and makes the last line break the
guarantees provided by the statically typed code.
In the first line, we assign an empty table to \texttt{person}.
In the second line, we assign the type of \texttt{person} to
\texttt{bogus}.
In the third and fourth lines we change the type of \texttt{person}.
In the fifth line we try to change the type of \texttt{bogus}
from \texttt{\{\}} to \texttt{\{"firstname":boolean\}},
but if we allow this type change we also allow changing the value
that is stored in \texttt{person.firstname}, regardless of its type.
Taking the changes individually looks fine, but the truth is that
aliasing makes one of them unsound.

We can create aliases to open variables, but they are always closed.
Any mutation in the original reference and in the aliased reference
is not a problem, as the type of the original reference can only
add new fields.
For mutable fields this means that the type of the field cannot
change after it is added to the type of the table.
In our previous example, the second line changes the type of
\texttt{person} from unique to open and makes the type of
\texttt{bogus} closed.

The location of the change also matters, as the next example shows:
\begin{verbatim}
    local person = { lastname = "Reed" }
    local function spoil ()
      person.firstname = nil          --> compile-time warning
    end
    person.firstname = "Lou"
    spoil()
    print("bye bye " .. person.firstname)
\end{verbatim}

In this example, the third line is unsound because this line allows
the programmer to change the type of \texttt{person} outside of the
scope that \texttt{person} was declared.
The call to \texttt{spoil} erases the field \texttt{firstname} from
\texttt{person}, and makes the last line break the guarantees
provided by the statically typed code.

@@

Typed Lua includes \emph{type aliases} for allowing programmers to
define their own data types.
Figure \ref{fig:aliases} shows the concrete syntax of the
\texttt{typealias} construct.

\begin{figure}[!ht]
\textbf{Type aliases}\\
\dstart
\begin{align*}
\textit{typealias} ::= & \;\; [\textbf{local}] \; \textbf{typealias} \; \textit{Name} \; \texttt{`='} \; \textit{type}
\end{align*}
\dend
\caption{The concrete syntax of Typed Lua type aliases}
\label{fig:aliases}
\end{figure}

As an example, the following declaration defines the type \texttt{Person}
as an alias to the table type \texttt{\{"firstname":string, "lastname":string\}}
in the remainder of the current scope:
\begin{verbatim}
    local typealias Person = {"firstname":string,
                              "lastname":string}
\end{verbatim}

Typed Lua also includes \emph{interfaces} as syntactic sugar to aliases
for table types that specify records, as writing table types can be unwieldy
when records get bigger and the types of record fields get more complicated.
Figure \ref{fig:interfaces} shows the concrete syntax of the
\texttt{interface} construct.
In Section \ref{sec:oop} we will show that we can also use \emph{interface} to
declare the type of classes, and that \emph{methodtype} is syntactic
sugar to express the type of methods.

\begin{figure}[!ht]
\textbf{Interfaces}\\
\dstart
\begin{align*}
\textit{interface} ::= & \;\; [\textbf{local}] \; \textbf{interface} \; \textit{typedec}\\
\textit{typedec} ::= & \;\; \textit{Name} \; { \; \textit{decitem} \; } \; \textbf{end}\\
\textit{decitem} ::= & \;\; \textit{idlist} \; \texttt{`:'} \; \textit{idtype}\\
\textit{idtype} ::= & \;\; \textit{type} \; | \; \textit{methodtype}\\
\textit{idlist} ::= & \;\; \textit{id} \; \{\texttt{`,'} \; \textit{id}\}\\
\textit{id} ::= & \;\; [\textbf{const}] \; \textit{Name}
\end{align*}
\dend
\caption{The concrete syntax of Typed Lua interfaces}
\label{fig:interfaces}
\end{figure}

As an example, we can use the following \texttt{interface} to
define the type \texttt{Person} from the previous example:
\begin{verbatim}
    local interface Person
      firstname:string
      lastname:string
    end
\end{verbatim}

After we define the type \texttt{Person}, we can use it in type annotations:
\begin{verbatim}
    local function byebye (person:Person):string
      return "Goodbye " .. person.firstname ..
             " " .. person.lastname
    end

@@

    print(greeter(user1))             --> Goodbye Lewis Reed 
    print(greeter(user2))             --> compile-time warning
    print(greeter(user3))             --> Goodbye Lou Reed
    print(greeter(user4))             --> compile-time warning
\end{verbatim}

@@

Even though the interface declaration may look redundant due to
the type alias declaration, it has a more convenient syntax
for declaring table types that express records.
For this reason, it worth having two different constructs;
one specifically for records and another for more general types.

The \texttt{interface} and \texttt{typealias} constructs also allow
the declaration of recursive types.
For instance, the following interface defines a type for singly-linked
lists of integers:
\begin{verbatim}
    local interface Element
      info:integer
      next:Element?
    end
\end{verbatim}

@@

    while l do
      print(l.info)
      l = l.next
    end
\end{verbatim}

Note that the type of \texttt{l} is \texttt{Element} in the
first line of the \texttt{while} body, but it is restored to
\texttt{Element|nil} in the second line of the \texttt{while} body
due to the assignment.
These rules allow this example to compile without any warnings,
and we discuss them in more detail in the next chapter.

As another example of recursive type, the following type alias
defines a type for the abstract syntax tree of a language of simple
arithmetic expressions:
\begin{verbatim}
    local typealias Exp = {"tag":"Number", 1:number}
                        | {"tag":"Add", 1:Exp, 2:Exp}
                        | {"tag":"Sub", 1:Exp, 2:Exp}
                        | {"tag":"Mul", 1:Exp, 2:Exp}
                        | {"tag":"Div", 1:Exp, 2:Exp}
\end{verbatim}
